#Reproducible Research Week 2 Project


=======================================

##First, load the data and necessary packages.

```{r}
library(ggplot2)
library(plyr)
data <- read.csv("activity.csv")
```

##Processing the data
```{r}
data$day <- weekdays(as.Date(data$date))
data$DateTime <- as.POSIXct(data$date, format="%Y-%m-%d")

#pulling data without NAs
cleandata <- data[!is.na(data$steps),]
```

##What is mean total number of steps taken per day?
### -Calculate total number of steps per day  
### - Plot a histogram of total number of steps taken each day  
### - Calculate the mean and the median
```{r}
#calculating total number of steps per day
sumTable <- aggregate(data$steps ~ data$date, FUN=sum,) 
colnames(sumTable)<- c("Date", "Steps")

#histogram of total number of steps per day
hist(sumTable$Steps, breaks=5, xlab="Steps", main="Total Steps per Day")

#mean and median total steps per day
mean(sumTable$Steps)
median(sumTable$Steps)
```

##What is the average daily activity pattern?
### - Time series plot of 5-minute interval (x-axis) and average number of steps taken, averaged across all days (y-axis)
### - Which 5 min interval, on avg across all the days, contains the maximum number of steps?
```{r}
library(plyr)
library(ggplot2)

#pulling data without NAs
cleandata <- data[!is.na(data$steps),]

#average number of steps per five-minute interval
intervalTable <- ddply(cleandata, .(interval), summarize, Avg = mean(steps))

#time-series plot of average steps per interval
ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab = "Average number of steps") + geom_line() + xlab("Interval") + ylab("Average number of steps") + ggtitle("Average number of steps per interval")

#maximum steps by interval
maxSteps <- max(intervalTable$Avg)

#which interval with max avg number of steps
intervalTable[intervalTable==maxSteps,1]
```

##Imputing missing values
### - Calculate and report total number of missing values in the dataset
### - Fill in the missing values. In this case, filled in mean values for that five-minute interval based on day of the week
### - Creat new data set that is the original data set but with missing data filled in
### - histogram of the total number of steps taken each day.  Calculate mean and median total nubmer of steps taken per day.
```{r}
missingvalues <- is.na(data$steps)
table(missingvalues)

avgTable <- ddply(cleandata, .(interval, day), summarize, Avg = mean(steps))

#create data set with all NAs for substitution
nadata <- data[is.na(data$steps),]

#merge NA data with average weekday interval for substitution
newdata1 <- merge(nadata, avgTable, by=c("interval", "day"))

#Reorder newly substituted data in the same format as original clean data set
newdata2 <- newdata1[,c(6,4,1,2,5)]
colnames(newdata2) <- c("steps", "date","interval","day", "DateTime")
#Merge NA averages and non NA data together
mergeData <- rbind(cleandata, newdata2)

sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN = sum,)
colnames(sumTable2) <- c("Date", "Steps")

hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")

mean(sumTable2$Steps)
median(sumTable2$Steps)
```
##Are there differences in activity patterns between weekdays and weekends?
### - Create new factor variable in the dataset with two levels - "weekday" and "weekend", indicating whether a given date is a weekday or weekend
### - Make panel plot containing a time series plot of five-minute interval (x-axis) and the average number of steps taken, average across all weekday days or weekend days (y-axis).
```{r}
#categorize based on days of week
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

library(lattice)

#summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

#plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="1", layout=c(1,2), main="Average steps per interval based on type of day", ylab="Average number of steps", xlab="Interval")
```